Analyse gender in music psychology from journal publication data

### 2025-09-18 manuscript/manuscript.qmd
Added text to conclusions and to the results

Taking a closer look at authorships, papers have explored certain aspects for such differences<to improve this last sentence to work way into this paragraph better>. One of these is different geographical regions. For example, men out-numbered women more so in high-income compared to low-income countries [@ioannidis2023gender]. In physics journals, female representation was lowest for Latin America and the Caribbean (16%), but highest for Africa (19%) [@son2022scientific]. In medical journals, women authors were least common in publications originating from South Korea, Japan, and Austria, but most common in South Africa and India [@bruck2023bibliometric]. Human development indicators and gender inequality indicators have also revealed systematic inequalities in authorship: Although female contribution to scientific output is lower in countries ranked in lowest quartile for human development and gender equity, this also occured in countries very high in gender equality; better equality seemed to occur in high and medium-developed countries [@sugimoto2015relationship].

@@

As the topics, approaches, and methods attract differential interest from men and women, it is important to recognise that fields of scholarship and academic disciplines vary considerably in terms of gender distribution [@huang2020]. For example, a large survey of authors by @gonzalez2020women showed that the hard sciences (N = 119,592) had the lowest prevalence of women (14.8%), whereas in the biological and social sciences (N = 262,122) the proportion was substantially higher (43.3%). In psychology, which is perhaps the closest benchmark for the present focus on music psychology, female authors accounted for 45.2% of the sample (N = 90,067).

To improve certain academic equity issues, it is worthwhile focusing on our discipline and to assess the equity in music psychology. The past efforts such as @anglada2019visualizing have already provided a current overview the publications in the music psychology field with suggestions of future directions to improve the field. @eerola2023transparency showcases how Open Science practices (e.g., preregistrations, sharing research materials, data, and analysis scripts) are relatively limited, and encourage authors in the field to start implementing such practices. @jakubowski2024 explores participant samples and musical stimuli used across a wide range of experiment to give an idea of how limited or generalizable the field is, which the view to see how – as a field – we need to diversify participant samples and stimuli to gain a more realistic understanding. In a similar way to these papers, the current paper aims to give a current overview of authorship and gender patterns in the field of music psychology.

@@

Our aim is to find out what is the gender distribution in the specialist journals of music psychology. What is the proportion of different types of authorships (single, first, coauthors, and last authorships) for men and women in the published papers in the last 25 years? Are there specific trends in terms of countries of the affiliations, topics, or time?

@@

For the analyses, we analysed gender disparities in authorship by comparing the frequency of these authorship types between male and female authors. Our analysis focused on the female author proportion (FAP) and odds ratios (ORs) comparing the likelihood of occupying each authorship position by gender.

@@

@@

To explore the potential differences within music psychology topics of choice between female and males, we analysed the keywords in the articles (15,316 in total, 6730 unique). To reduce variant keywords, we manually simplified the variants (e.g., "arts in health", "healthy music use", "health musicking" were converted to "music health" and "music cognition" to "cognition" and "music perception" to "perception") and eliminated single use of the term "music". We aggregated the counts of keywords across all authors in the articles and Figure 4 shows the odds ratios for female authors to be associated (in any author position) with the 40 most frequently used keywords. The keywords signifying related to children, development, emotion regulation, well-being, music performance anxiety, motivation, expertise, music listening, mental health, and music therapy show significantly higher odds ratios (OR $\ge$ 1.50) for females in comparison to overall gender ratios in the data. At the opposite end of this continuum, topics such as music analysis, harmony, machine learning, computation, cognition, and performance expression show the opposite trend, suggesting that the articles with these keywords tend to have overall more prevalent male authorships (female OR $\le$ 0.75).

@@

There were substantial gender differences in citation patterns. One notable finding is that first-authored articles by women were significantly more likely to be published open access, with an odds ratio of 1.49, a result that warrants further exploration. This effect does not appear to be tied to specific journals. Although the odds ratio for women publishing open access is highest in _Music and Science_, the broader trend suggests a temporal shift: since around 2010, women have increasingly overtaken men in publishing open access. The data does not provide any reasons for this. While past surveys among UK academics have not indicated substantial differences in experiences with open access [@zhu2017support], financial and other inequalities in emerging countries have been suggested to limit women's capacity to publish in open access journals [@vuong2021adopting].

@@

The keyword analysis of the articles indicated thematic differences in the topics studied by women and men. Research areas related to children, well-being, health, and mental health were more frequently associated with women. This pattern is consistent with observations in psychology more broadly, where women are more strongly represented in sub-disciplines that focus on care and nurturing [@gonzalez2020women]. The higher likelihood of women studying topics related to singing may be connected to prevailing gender stereotypes, or to the early negative perceptions of singing among boys [@Warzecha2013] and men [@palkki2015gender]. In contrast, topics related to cognition and computation were more often associated with men, which reflects a wider trend in psychology [@gonzalez2020women].

[@ashmos2011gendering; @sebo2020gender; @thelwall2019gender], In a similar trend in computational linguistics, woman are more likely to publish on speech, social and conversational topics than males, who are more likely to publish in formal mathematical, syntax, and semantic approaches [@vogel2012he].

@@

The analysis of gender distributions in the five most prominent music psychology journals over the last 25 years presented a rich and evolving history of gender distribution in various authorship roles. While overall women hold smaller number of authorship overall (`{r} round(nrow(dplyr::filter(df,Gender=='female'))/nrow(df),3)*100`%), they are more likely to hold first author positions in coauthored publications but less likely likely to hold last author positions than men. For solo publications and being a coauthor, there are no significant gender differences.

There was small but significant effect of gender on citations that favoured men when the citations were aggregated all authors in the papers. There was no, however, differences in citations for female first or last  authors. The aggregated results are in line with previous findings in psychology that favored men [@gonzalez2020women]. We also observed that female first authored articles tend to be more likely to be published as open access as male first authors. The past surveys of experiences and attitudes towards open access publishing has revealed gender differences typically in the opposite direction [@zhu2017support].

While the present analysis cannot determine the underlying causes of the observed trends, we can point to several factors that may help explain why this discipline appears to be more equitably distributed by gender than, for example, psychology.

(1) Female role models (past and present) exist in music psychology; many of the founding members of the societies have been women (e.g., Diana Deutsch and Edward Carterette for ICMPC in 1989, Irène Deliège for ESCOM in 1991), to a more recent ICMPC and ESCOM presidents? Editor-in-chiefs of the journals?).

(2) Taking sub-disciplinary differences in psychology as an indicator of topic areas more often favoured by women, @gonzalez2020women observed that “nurturing” disciplines such as education and developmental psychology had higher proportions of female authors. In music psychology, however, a similar explanation may not apply. Keywords typically associated with nurturing domains (e.g., education, development, social, therapy, well-being, mental health, emotion regulation) account for only 17.9% of the total keyword counts among the top 40 keywords.

- It would have been interesting to analyse the precarity of career options as judged from the years of authorship attributions [@lundine2018]

- Current findings present a more optimistic outlook of female authors in comparison what the prevalence of female academics or professors in academia are [@bhopal2021competing;@harris2025gender;@Filandri03082021]. While we have comparable accounts from psychology, musicology and other disciplines of music research (popular music, ethnomusicology) tends to have less equal division gender; a recent analysis of the last 25 years of articles in 15 musicology journals amounting to 3,515 articles suggests that female authors are far more infrequent (34.8% [33.3%-36.3%]) than males (65.2% [63.7%-66.7%]) [@eerola2026inprep].

- Senior/last authorship position and why such a large inbalance exists.

>>>>

Past studies have explored the gender distribution across a range of elements that could illuminate the reasons for the discrepancies. One of these is geographical regions. For example, men out-numbered women more so in high-income compared to low-income countries [@ioannidis2023gender]. In physics journals, female representation was lowest for Latin America and the Caribbean (16%), but highest for Africa (19%) [@son2022scientific]. In medical journals, women authors were least common in publications originating from South Korea, Japan, and Austria, but most common in South Africa and India [@bruck2023bibliometric]. Human development indicators and gender inequality indicators have also revealed systematic inequalities in authorship: Although female contribution to scientific output is lower in countries ranked in lowest quartile for human development and gender equity, this also occured in countries very high in gender equality; better equality seemed to occur in high and medium-developed countries [@sugimoto2015relationship].

@@

As the topics, approaches, and methods attract differential interest from men and women, it is important to recognise that fields of scholarship and academic disciplines vary considerably in terms of gender distribution [@huang2020]. For example, a large survey of authors by @gonzalez2020women showed that the hard sciences (N = 119,592) had the lowest prevalence of women (14.8%), whereas in the biological and social sciences (N = 262,122) the proportion was substantially higher (43.3%), but not equal. In psychology, which is perhaps the closest benchmark for the present focus on music psychology, female authors accounted for 45.2% of the sample (N = 90,067).

To improve academic equity issues, it is worthwhile focusing on our discipline and to assess the equity in music psychology. The past efforts such as @anglada2019visualizing have already provided a current overview the publications in the music psychology field with suggestions of future directions to improve the field. @eerola2023transparency showcases how Open Science practices (e.g., preregistrations, sharing research materials, data, and analysis scripts) are relatively limited, and encourage authors in the field to start implementing such practices. @jakubowski2024 explores participant samples and musical stimuli used across a wide range of experiment to give an idea of how limited or generalizable the field is. In a similar way to these papers, the current paper aims to give a current overview of gender patterns in authorship in the field of music psychology.

@@

Our aim is to find out what is the gender distribution in the specialist journals of music psychology. More specifically, we want to identify what is the proportion of different types of authorships (single, first, coauthors, and last authorships) for men and women in the published papers in the last 25 years? Are there specific patterns in the authorship distributions across countries, topics of the studies, or temporal trends?

@@

For the analyses, we analysed gender disparities in authorship by comparing the frequency of these authorship types between male and female authors. Our analysis focused on the female author proportion (FAP) and utilised odds ratios (ORs) to compare the likelihood of occupying each authorship position by gender.

@@

Panel A in Figure 3 shows a small number of countries, such as Australia, China, and Canada, with a high odds ratio for female authorship. At the other end of the spectrum, Japan, Belgium, Austria, Sweden, and Turkey all have significantly lower odds ratios for female authors. There is no simple rationale immediately apparent (e.g., country size, membership in a particular political bloc, gender equality index, or domination of articles by a single eminent scholar from one country) that would explain these differences, but this representation provides some starting points for discussion and may highlight the impact of academic career structures and principles implemented in these countries. Panel B shows data aggregated by continent, where Africa has meagre representation (52 authors), Oceania consists only of Australia and New Zealand, and the Americas are represented almost entirely by the US and Canada. While this level of presentation may be too coarse for detailed analysis, it does offer a broad summary of the current situation.

@@

To explore the potential differences within music psychology topics of choice between female and males, we analysed the keywords in the articles (15,316 in total, 6,730 unique). To reduce variant keywords, we manually simplified the variants (e.g., "arts in health", "healthy music use", "health musicking" were converted to "music health" and "music cognition" to "cognition" and "music perception" to "perception") and eliminated single use of the term "music". We aggregated the counts of keywords across all authors in the articles and Figure 4 shows the odds ratios for female authors to be associated (in any author position) with the 40 most frequently used keywords. The keywords signifying related to children, development, emotion regulation, well-being, music performance anxiety, motivation, expertise, music listening, mental health, and music therapy show significantly higher odds ratios (OR $\ge$ 1.50) for females in comparison to overall gender ratios in the data. At the opposite end of this continuum, topics such as music analysis, harmony, machine learning, computation, cognition, and performance expression show the opposite trend, suggesting that the articles with these keywords tend to have overall more prevalent male authorships (female OR $\le$ 0.75). There is much nuance and detail in the data that would warrant separate analyses; however, the purpose here is simply to provide reference points for comparing gender distributions of authorship with other disciplines, where thematic analyses have already been conducted (e.g., in sub-disciplines of psychology).

@@

There were no substantial gender differences in the citation patterns. However, one notable observation is that first-authored articles by women were significantly more likely to be published open access, with an odds ratio of 1.49, a result that warrants further exploration. This effect does not appear to be tied to specific journals. Although the odds ratio for women publishing open access is highest in _Music and Science_, the broader trend suggests a temporal shift: since around 2010, women have increasingly overtaken men in publishing open access. The data does not provide any direct reasons for this, although the legislation in few of the European nations (United Kingdom, EU) require open access publishing and when women have higher odds ratio for being first authors, this might be due to the high number of female from these countries. While past surveys among, for instance, UK academics have not indicated substantial differences in experiences with open access [@zhu2017support], financial and other inequalities in emerging countries have been suggested to limit women's capacity to publish in open access journals [@vuong2021adopting]. But nothing in the extant literature would point out to the reverse pattern, which requires further investigation about the reasons for adopting open access publishing models.

@@

The keyword analysis of the articles indicated thematic differences in the topics studied by women and men. Research areas related to children, well-being, health, and mental health were more frequently associated with women. This pattern is consistent with observations in psychology more broadly, where women are more strongly represented in sub-disciplines that focus on care and nurturing [@gonzalez2020women]. The higher likelihood of women studying topics related to singing may be connected to prevailing gender stereotypes, or to the early negative perceptions of singing among boys [@Warzecha2013] and men [@palkki2015gender]. In contrast, topics related to cognition and computation were more often associated with men, which reflects a wider trend in psychology [@gonzalez2020women]. In a similar trend in computational linguistics, woman are more likely to publish on speech, social and conversational topics than males, who are more likely to publish in formal mathematical, syntax, and semantic approaches [@vogel2012he].

<!-- I did not know what these references signaled, so temporarily disabled them: -->
<!-- [@ashmos2011gendering; @sebo2020gender; @thelwall2019gender], -->

@@

The analysis of gender distributions in the five most prominent music psychology journals over the last 25 years presented a rich and evolving history of gender distribution in various authorship roles. While overall women hold smaller number of authorship overall (`{r} round(nrow(dplyr::filter(df,Gender=='female'))/nrow(df),3)*100`%), they are significantly more likely to hold first author positions in coauthored publications but significantly less likely likely to hold last author positions than men. For solo publications and being a coauthor, there are no significant gender differences.

There was small but significant effect of gender on citations that favoured men when the citations were aggregated all authors in the papers. There was no, however, differences in citations for female first or last authors. The aggregated results are in line with previous findings in psychology that favored men [@gonzalez2020women]. While the present analysis cannot determine the underlying causes of the observed trends, we can point to potential factors that may help explain why this discipline appears to be more equitably distributed by gender for the first authors than, for example, psychology. Female role models (past and present) exist in music psychology; many of the founding members of the societies have been women (e.g., Diana Deutsch for ICMPC in 1989, Irène Deliège for ESCOM in 1991, and similarly many founding editors of the journals have been women (e.g., Diana Deutsch for _Music Perception_ 1983-1995, and Irène Deliège for Musicae Scientiae 1997-2009). However, this does not fully explain the rise of female first authors over the past 15 years, nor is it consistent with the reverse pattern observed in the last authorship position. A cautious interpretation is that the increase in first authorship may signal the discipline becoming more gender-balanced, but this shift has not yet extended to last authorship positions, which still reflect either career longevity or barriers to women attaining senior academic positions.

While the analysis of gender of the authorships across the discipline-specific journal is relatively simple summary in the present study, there are more elaborate ways into obtaining insights into the gendered career progression in this disciple; for instance, future work could analyse the precarity of career options as judged from the years of authorship attributions [@lundine2018]. One could also analyse the last author positions in more detail using direct contacts to the authors (interviews and surveys) and try to pin down the factors contributing to the barriers experienced by women to take up senior author positions.

Current findings present a more optimistic outlook of female authors in comparison what the prevalence of female academics or professors in academia are [@bhopal2021competing;@harris2025gender;@Filandri03082021]. While the results are comparable to psychology in general, musicology (and the closely associated disciplines such as music theory, popular music and ethnomusicology) tend to have less equal division of gender in overall authorship; a recent analysis of the last 25 years of articles in 15 most eminent musicology journals amounting to 3,515 articles suggests that female authors are far more infrequent (34.8% [33.3%-36.3%]) than males (65.2% [63.7%-66.7%]) [@eerola2026inprep]. The present findings should be used in conjuction of ...
